A weakly structured stem for human origins in Africa

### 2023-02-20 paper.tex
minor typo fixes

We quantify migration rates among populations after their divergence
$\sim120$ka. Prior to agripastoralist expansion $5$ka, migration between the
ancestors of the Nama and other groups is an order of magnitude weaker than
that observed between between western and eastern Africans (Table
\ref{tab:migration-rates}). All models infer relatively high gene flow between
eastern and western Africa ($m\approx2\times10^{-4}$, the constant proportion
of immigrant lineages per generation over $60$ka). We further find that
Back-to-Africa gene flow at the beginning of the Holocene primarily affected
the ancestors of the Ethiopian agricultural populations
\citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
estimated to be approximately $65\%$. We observe significant gene flow from the
Amhara and Oromo into the Nama, a signal which is likely a proxy for the
movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a $25\%$ ancestry
contribution 2,000 years ago. While this gene flow is not apparent from the
\texttt{ADMIXTURE} plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
into the Khoe-San component which has drifted appreciably from its ancestral
eastern African source. Colonial period admixture from Europeans into the Nama
was estimated at $15\%$, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups.  We observe a
sharp bottleneck in Stem 1 down to $N_e=100$ after the split of the Neanderthal
branch. This represents the lower bound allowed in our optimization (i.e. an
$N_e$ of 100), although the size of this bottleneck is poorly constrained (95\%
confidence interval 100--851). After a long period of exchange with Stem 2,
Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S'' 478ka. The timing of
this divergence was also poorly constrained ($95\%$ CI $276-478$ka).  These
populations evolve independently until $119$ka ($101-125$ka) when Stem 1S and Stem
2 combine to form the ancestors of the Nama, with proportions 30\% and 70\%,
respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions (50\%
each) to form the ancestors of the western Africans and eastern Africans (and
thus also all individuals who later disperse during the Out-of-Africa event).
Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
replacing 19\% (18--21\%) of their population $25$ka ($22-26$ka). The later Stem 2
contribution to the western African Mende resulted in better model fits
($\Delta LL \approx 60,000$). Speculatively, this may indicate that an
ancestral Stem 2 population occupied western or central Africa, broadly
speaking. The differing proportions in the Nama and eastern Africans may also
indicate geographic separation of Stem 1S in southern Africa and Stem 1E in
eastern Africa.

To assess robustness of the inferred models to analysis and reference
population choices, SI Sections \ref{sec:validation_strategy}
and \ref{sec:validation_results} present re-analyses changing the European and
West African populations, as well as the recombination maps, filtering
strategies and parameter optimization strategies. While we find some
differences in the inferred parameters (see in particular Sections
\ref{sec:replacing} and \ref{sec:change-dates}), the best fit model across all
reanalyses are quantitatively consistent.

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred
models. Many previous studies have found a reduction of coalescence rates
between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
could be due to either an increase in population size or to ancestral
population structure during the Middle Pleistocene \citep{Mazet2016-wn}.  All
models, including the single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between $100$ka--$1$Ma
(Figure~\ref{fig:supp-iicr-sim} and SI
Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
increase in $N_e$ during that period, the best-fit model recapitulate this
pattern without corresponding population size changes.

@@

Our inferred models paint a consistent picture of the Middle to Late
Pleistocene as a critical period of change, assuming that estimates from the
recombination clock accurately relate to geological chronologies (SI
Section~\ref{sec:supp-mutation}). During the late Middle Pleistocene, the
multiple merger model indicates three major stem lineages in Africa,
tentatively assigned to southern (Stem 1S), eastern (Stem 1E) and
western/central Africa (Stem 2). Geographic association was informed by the
present population location with the greatest ancestry contribution from each
stem. For example, Stem 1S contributes 70\% to the ancestral formation of the
Khoe-San. The extent of the isolation $\>400$ka between Stem 1S, Stem 1E and Stem
2 suggests that these stems were not proximate to each other. While the length
of isolation among the stems is variable across fits, models with a period of
divergence, isolation and then a merger event (i.e. a ``reticulation'')
out-performed models with bifurcating divergence and continuous gene flow.

@@

We also find that Stem 2 continued to contribute to western Africans during the
Last Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}). Such an
interpretation is reinforced by the differential migration rates between
regions, i.e. the gene flow from Stem 2 to western Africans is estimated to be
five times that of the rate to eastern Africans during this period. We
performed a variety of validation tests to explore sensitivity of our
assumptions, including relaxing fixed parameters (SI
Section~\ref{sec:validation_strategy}). Most validation tests resulted in
parameters similar to the models discussed above. However, one exception was
the inferred Out-of-Africa and eastern/western African divergences which were
10--15ka younger than our fixed parameters. The
complexity of the OOA expansion warrants the inclusion of additional Eurasian populations 
and more detailed modeling. However, other features
of our inferred models are weakly affected by this analysis choice.

>>>>

We quantify migration rates among populations after their divergence
$\sim120$ka. Prior to agripastoralist expansion $5$ka, migration between the
ancestors of the Nama and other groups is an order of magnitude weaker than
that observed between western and eastern Africans (Table
\ref{tab:migration-rates}). All models infer relatively high gene flow between
eastern and western Africa ($m\approx2\times10^{-4}$, the constant proportion
of immigrant lineages per generation over $60$ka). We further find that
Back-to-Africa gene flow at the beginning of the Holocene primarily affected
the ancestors of the Ethiopian agricultural populations
\citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
estimated to be approximately $65\%$. We observe significant gene flow from the
Amhara and Oromo into the Nama, a signal which is likely a proxy for the
movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a $25\%$ ancestry
contribution 2,000 years ago. While this gene flow is not apparent from the
\texttt{ADMIXTURE} plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
into the Khoe-San component which has drifted appreciably from its ancestral
eastern African source. Colonial period admixture from Europeans into the Nama
was estimated at $15\%$, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups.  We observe a
sharp bottleneck in Stem 1 down to $N_e=100$ after the split of the Neanderthal
branch. This represents the lower bound allowed in our optimization (i.e., an
$N_e$ of 100), although the size of this bottleneck is poorly constrained (95\%
confidence interval 100--851). After a long period of exchange with Stem 2,
Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S'' 478ka. The timing of
this divergence was also poorly constrained ($95\%$ CI $276-478$ka).  These
populations evolve independently until $119$ka ($101-125$ka) when Stem 1S and Stem
2 combine to form the ancestors of the Nama, with proportions 30\% and 70\%,
respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions (50\%
each) to form the ancestors of the western Africans and eastern Africans (and
thus also all individuals who later disperse during the Out-of-Africa event).
Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
replacing 19\% (18--21\%) of their population $25$ka ($22-26$ka). The later Stem 2
contribution to the western African Mende resulted in better model fits
($\Delta LL \approx 60,000$). Speculatively, this may indicate that an
ancestral Stem 2 population occupied western or central Africa, broadly
speaking. The differing proportions in the Nama and eastern Africans may also
indicate geographic separation of Stem 1S in southern Africa and Stem 1E in
eastern Africa.

To assess robustness of the inferred models to analysis and reference
population choices, SI Sections \ref{sec:validation_strategy}
and \ref{sec:validation_results} present re-analyses changing the European and
West African populations, as well as the recombination maps, filtering
strategies and parameter optimization strategies. While we find some
differences in the inferred parameters (see SI sections
\ref{sec:replacing} and \ref{sec:change-dates}), the best fit models across all
reanalyses are quantitatively consistent.

@@

We used \texttt{Relate} \citep{Speidel2019-nj} to infer the distribution of
coalescence rates over time in real data and data simulated from our inferred
models. Many previous studies have found a reduction of coalescence rates
between $1000$ka and $100$ka in humans, and thus inferred an increase in $N_e$
during the same period \citep{Li2011-le}.  This increase in inferred $N_e$
could be due to either an increase in population size or to ancestral
population structure during the Middle Pleistocene \citep{Mazet2016-wn}.  All
models, including the single-origin model, recapitulate an inferred ancestral
increase in $N_e$ between $100$ka--$1$Ma
(Figure~\ref{fig:supp-iicr-sim} and SI
Section~\ref{inferred_relate}). Whereas the single-origin model achieves this by an
increase in $N_e$ during that period, the best-fit models recapitulate this
pattern without corresponding population size changes.

@@

Our inferred models paint a consistent picture of the Middle to Late
Pleistocene as a critical period of change, assuming that estimates from the
recombination clock accurately relate to geological chronologies (SI
Section~\ref{sec:supp-mutation}). During the late Middle Pleistocene, the
multiple merger model indicates three major stem lineages in Africa,
tentatively assigned to southern (Stem 1S), eastern (Stem 1E) and
western/central Africa (Stem 2). Geographic association was informed by the
present population location with the greatest ancestry contribution from each
stem. For example, Stem 1S contributes 70\% to the ancestral formation of the
Khoe-San. The extent of the isolation $\>400$ka between Stem 1S, Stem 1E and Stem
2 suggests that these stems were not proximate to each other. While the length
of isolation among the stems is variable across fits, models with a period of
divergence, isolation and then a merger event (i.e., a ``reticulation'')
out-performed models with bifurcating divergence and continuous gene flow.

@@

We also find that Stem 2 continued to contribute to western Africans during the
Last Glacial Period, indicative that this gene flow likely occurred in
western/central Africa (Table~\ref{tab:migration-rates}). Such an
interpretation is reinforced by the differential migration rates between
regions, i.e., the gene flow from Stem 2 to western Africans is estimated to be
five times that of the rate to eastern Africans during this period. We
performed a variety of validation tests to explore sensitivity of our
assumptions, including relaxing fixed parameters (SI
Section~\ref{sec:validation_strategy}). Most validation tests resulted in
parameters similar to the models discussed above. However, one exception was
the inferred Out-of-Africa and eastern/western African divergences which were
10--15ka younger than our fixed parameters. The
complexity of the OOA expansion warrants the inclusion of additional Eurasian populations 
and more detailed modeling. However, other features
of our inferred models are weakly affected by this analysis choice.
